---
title: "tRee2MillCost"
author: "Demetrios Gatziolis"
date: "`r Sys.Date()`"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{tRee2MillCost}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse=TRUE,
  comment="#>",
  out.width="90%" )
```

```{r setup}
library(tRee2MillCost)
```
## Package description

`tRee2MillCost` is an `R` package that calculates travel cost across a
rasterized representation of a road network between origin (from) and
destination (to) locations. The cost can be calculated as minimal travel time or
travel distance. As the name suggests, the package was originally developed for
the [BioSum](http://biosum.info) decision support framework to estimate the cost
of transporting woody material from forests to processing locations such as
mills.

Road networks are typically represented as vector line layers obtained from
various sources. These layers tend to use different attribute schemas and often
lack consistent topology. This is common for forest, rural, and urban road
networks alike. The absence of explicit topology, often referred to as a
‘spaghetti’ data model, impedes optimal route calculations. Developing a
topologically clean vector network is labor-intensive and often impractical.
This package addresses the limitation by converting a composite assembled from
multiple vector networks to a raster representation.

A composite vector network ingested to produce the raster representation does
not require road attributes to produce cost estimates based on travel distance.
For cost estimates based on travel time, all segments in the composite vector
network must have an attribute representing the modeled travel speed along that
segment—for example, the expected speed of log trucks. Legal speed limits are
one way to represent travel speed, but rarely included in publicly available
vector road layers. Users therefore must estimate travel speeds from auxiliary
information, and appropriate values for a given road class may vary
geographically. The vector road layer must be attributed with travel speed
values before running package functions. Raster cells containing more than one
road segment, or part of a segment, inherit the highest travel speed value among
them.

The cell values of the raster road layer are then used to determine travel
distance from each road cell to any of its 8 immediate neighbors that are also
members of the road network, analogous to a king’s move in chess. Unlike the
chess rook’s move rule that only considers orthogonal neighbors, the king’s rule
also allows for travel to a diagonal neighbor at a `cellSize`$\sqrt{2}$
distance, a setting that yields more realistic road representations and
efficient travel calculations. For cost estimates in travel time distances are
converted to time using the mean travel speed assigned to each of the two 
connected cells. Road cells are then represented as graph nodes, enabling
least-travel-distance or least-travel-time route calculations with the
`cppRouting` package.

Rasters representing road networks are in essence sparse matrices. Route
calculations between an origin and a destination require both locations to
coincide with road cells. The package workflow examines travel origins and
destinations and, when necessary, moves them to the nearest road-network cell in
2-dimensional space; the displacement distance is recorded and, for modeling the
cost of moving trees from an origin location to a roadside landing, can serve as
a proxy for yarding distance. The minimum travel distance or time is then
calculated for every repositioned origin-destination pair. The resulting cost
matrix can be used to calculate destination- or origin-level summaries and
metrics. The package can also extract optimal (minimum travel distance or time)
routes between on-road cells and isochrones, defined as portions of the network
reachable from a road cell within a specified travel-distance or travel-time
threshold.

Consistent with the GDAL spatial-data model, all rasters created by the package
are saved to the disk. Speed values are converted to 8-bit (1 byte) integers,
with 255 reserved for background (no network) cells. Speed values are expected
to be expressed in miles per hour. Computationally intensive operations not
handled directly by the `cppRouting` are implemented in embedded `Rcpp`.
`tRee2MillCost` can process road rasters with more than 2^31^ cells and
thousands of origin and destination locations. However, excessively fine
resolutions (< 5m) do not improve the information content of the network
representation and may leave numerous road segments disconnected from the rest
of the network, thereby reintroducing the connectivity problem that
rasterization is intended to alleviate.

Once the raster representation is constructed, it is examined for connectivity.
Road cells with no immediate neighbors or groups of cells that are otherwise
disconnected from the rest of the network are detected, exported as GeoPackage
point layers, and removed from the network. Users can examine the spatial
distribution of disconnected segments and, when appropriate, manually edit the
vector network to connect them. After any edits, the rasterization and
subsequent processing steps must be repeated. 

## Main functions

- `calculateCost` : calculate travel cost in time or distance between origin and
destination locations
- `connectivity` : determine whether points lie on road network cells
- `createGraph` : create a routing graph from a road-network raster
- `movePt2RdSegment` : move origin or destination points to road network cells
- `prepareGraph` : create a cell-connectivity data frame from a raster
- `rasterizeRoads` : rasterize a vector line file

## Auxiliary functions

- `backgroundCellValue` : set raster background cell values to NA
- `cellFromGraphNode` : get the road-network cell ID corresponding to a graph
node
- `checkInputs` : validate processing parameters and input data
- `costSummary` : summarize travel time or distance between origins and
destinations
- `fromToPaths` : delineate routes between sets of origin and destination
locations
- `getIsochrone` : identify and export the portion of the road network
accessible from a node within a specified travel time or distance threshold
- `graphNodeFromXY` : determine the graph node ID corresponding to a coordinate
- `maskCells` : set selected raster-cell values to NA
- `setCellValueToInteger` : efficiently convert a numeric raster to integer type

## Workflow example

```{r, message=FALSE, warning=FALSE}
library( tRee2MillCost )
library( terra )
```

```{r, results='hide'}
# All spatial data must share the same projected coordinate system with
# linear unit either in meters or feet. Unprojected or geographic
# coordinate systems (longitude/latitude) cannot be used

# Path to the vector layer of the road network
rdPath <- system.file( "extdata", "roads.gpkg", package="tRee2MillCost" )
# Path to the layer containing travel origin ('from') locations / points,
# forest inventory plots in this example
fromPath <- system.file( "extdata", "sample.gpkg", package="tRee2MillCost" )
# Path to the layer containing destination travel ('to') locations / points,
# woody material processing facilities (mills) in this example
toPath <- system.file( "extdata", "mills.gpkg", package="tRee2MillCost" )

# Set the names of fields (attributes) in rdPath, fromPath, and toPath
# layers. All must exist. rdField values should be numeric
rdField   <- "MPH"
fromField <- "sampleID"
toField   <- "COMPANY_NAME"

# Set the resolution of the road network raster in the linear units of the
# projection used. In this example, a raster resolution of 20 implies meters
# because rdPath, fromPath, and toPath are in UTM zone 10N projection
rasterResolution <- 20

# Specify if travel cost will be calculated in time (costType set to 1) or
# distance (costType set to 2)
costType       <- 1 # travel calculations for fastest travel time

# Specify outputs
roadRasterPath <- "road_net.tif"
movedFromPath  <- "moved_plots.gpkg"
movedToPath    <- "moved_mills.gpkg"
costCSVpath    <- "cost_matrix.csv"

# Determine validity of inputs. Evaluate any warnings and decide if you want
# to proceed
checkInputs( rdPath, rdField, toPath, toField, fromPath, fromField,
             rasterResolution, roadRasterPath,
             movedToPath, movedFromPath,
             costType,
             costCSVpath )

# Convert the vector representation of the road network to raster
# Specify rdField so that future cost computations will be travel time-based
# If rdField is omitted, cost computations will be distance-based
rasterizeRoads( rdPath, rasterResolution, roadRasterPath, rdField )

# Create the graph
# prepareGraph() is called by createGraph() internally, it is not expected to
# be deployed independently
graph <- createGraph( roadRasterPath, costType )
saveRDS( graph, "graph.rds")

# Determine if all plots are on-network locations
connectivity( roadRasterPath, fromPath )
# Determine if all mills are on-network locations
connectivity( roadRasterPath, toPath )

# Since they are not, move them to on-network locations
movePt2RdSegment( roadRasterPath, fromPath, movedFromPath )
movePt2RdSegment( roadRasterPath, toPath,   movedToPath )

# Confirm that all plots and mills are now on the network
# Note that the point layers are now movedFromPath and movedToPath
connectivity( roadRasterPath, movedFromPath )
# Determine if all mills are on-network locations
connectivity( roadRasterPath, movedToPath )

# Calculate travel cost
cost.df <- calculateCost( graph,
                          roadRasterPath,
                          movedFromPath, fromField,
                          movedToPath, toField )

# Save the cost matrix to the disk
write.csv( cost.df, costCSVpath, row.names=FALSE )
```

```{r}
# Summarize travel time for plots that moved from their original location to
# the road network by no more than 500m and then routed on-network to the
# closest mill in no more than 30 minutes (0.5 hours)
cost.list <- costSummary( cost.df, movedFromPath, fromField,
                          maxMoveDistance=500, travelThreshold=0.5 )
print( cost.list[[1]] )
print( cost.list[[2]] )
```

```{r, iso, fig.width=7.5, fig.height=7.5}
# Create and plot a 30-minute isochrone raster for one of the mills
millName <- "Mount Vermion Industries"
sel <- data.frame( vect( movedToPath ) )[, toField] == millName
mill <- vect( movedToPath )[sel]
millCellID <- cellFromXY( rast( "road_net.tif" ), crds( mill ) )
getIsochrone( graph, millCellID, 0.5,  millName )
iso.r <- rast( "isochrone_0.5h_Mount Vermion Industries.tif" )
pal <- colorRampPalette( c("#332288", "#88CCEE", "#44AA99", "#117733",
                           "#999933", "#DDCC77", "#CC6677", "#882255",
                           "#AA4499", "#DDDDDD") )
par( mar=c( 1.5, 1, 4, 3 ) )
plot( iso.r,
      maxcell=1E12, background="black", mar=NA,
      smooth=FALSE, col=pal( 256 ),
      plg=list( title="Travel\nTime\n(hours)" ),
      main=paste("30-minute isochrone from", millName, "\n") )
points( mill, pch=16, cex=2, col="white" )
```

```{r, paths, fig.width=9, fig.height=9 }
# Determine and plot the routes for all plots to a mill
plots <- vect( movedFromPath )
plotIDs <- data.frame( plots[, fromField] )[,1]
millID <- as.character( data.frame( mill[, toField] ) )
paths <- fromToPaths( graph, "road_net.tif",
                      "moved_plots.gpkg", fromField, plotIDs,
                      "moved_mills.gpkg", toField, millID )
par( mar=c( 1, 1, 0.1, 0.1 ) )
terra::plot( vect( ext(paths) + 500 ) )
terra::lines( crop( vect(rdPath), vect(ext(paths) + 500) ),
              lwd=0.5, col="gray")
terra::lines( paths )
legend( x = 503000, y=4896000,
        legend=c("Mill", "Plots") ,col=c("blue", "red"),
        cex=1.5, pch=16, bg="beige" )
# 'travel time' in title(), as costType is set to 1
title( paste( "Shortest travel time routes from plots to", millID ),
       line=-1.5, cex.main=1.5 )
terra::points( plots, pch=16, col="red",  cex=1.25 )
terra::points( mill,  pch=16, col="blue", cex=1.25 )
```
